% ============================================================
% Conclusions (target: Computers & Security)
% ============================================================
\section{Conclusions}
\label{sec:conclusion}

Modern phishing detection is effective in aggregate, yet a residual fraction --- spear phishing
and~BEC --- remains the most costly and~almost invisible to content filters. We have shown that
this is not an accident but a structural property: a targeted attack succeeds \emph{precisely
because} it respects the victim's domain knowledge, and~therefore the maliciousness signal lies
not in the content but in the \emph{inconsistency} with the graph of that knowledge and~in the \emph{dynamics} of its violation.

\paragraph{Answers to the research questions.}
\begin{description}[noitemsep,topsep=2pt]
  \item[Does the domain knowledge graph detect attacks invisible to content?] \emph{Yes, conditionally.} The cross-layer
  inconsistency invariant beats a single contact graph and~eliminates false alarms from
  legitimate new senders (Section~\ref{sec:res-inv1}), provided there is a shared, observable domain-knowledge
  edge.
  \item[Does cascade dynamics detect a compromised account?] \emph{Yes.} The temporal invariant removes the blindness
  of static layers and~--- unlike a volume-based detector --- \emph{does not collapse} against a stealthy
  attacker (Section~\ref{sec:res-inv2}); the advantage is causally temporal.
  \item[What is the organizational susceptibility?] \emph{Measurable.} It is governed by the \emph{relation}, not
  rank: cold external impersonation leaves a margin of $\mathrm{Susc}\!\approx\!0.96$ at every
  level, whereas in the lateral channel it is the \emph{periphery} of the organization that is most exposed, not
  the executives (Section~\ref{sec:susceptibility}).
  \item[What does it mean for the method to ``work''?] We define effectiveness operationally and~conditionally ---
  through sensitivity at a fixed FPR under leak controls and~a predictiveness probe ---
  with an explicitly stated floor (cold impersonation) and~an honest threshold (Section~\ref{sec:res-rigor}).
\end{description}

\paragraph{Contributions.}
\begin{enumerate}[noitemsep,topsep=2pt]
  \item \textbf{The concept and~formalization of the domain knowledge graph} together with the thesis that spear phishing is its
  violation --- uniformly for lateral/BEC and~external impersonation.
  \item \textbf{Two complementary detection invariants} (cross-layer inconsistency
  and~cascade dynamics), with causal controls and~validation on three real communication graphs.
  \item \textbf{A measured organizational susceptibility metric} revealing that the residual margin
  is governed by the relation, and~that in the lateral channel the periphery is more exposed than the top.
  \item \textbf{A leak-aware evaluation methodology} and~an operational definition of spear phishing detection
  effectiveness.
\end{enumerate}

\paragraph{Limitations (summary).} The maliciousness is synthetic (we validate the mechanism, not
the detection of real intrusions); the susceptibility measurement inherits the assumptions of the event model; scale
and~the operational threshold remain limitations --- details in Section~\ref{sec:discussion}.

The contribution is not a finished product but \emph{a concept, a method, and~measurement rigor}. Spear phishing is
a violation of domain knowledge; the graph of that knowledge makes it --- and~the susceptibility to it --- \emph{measurable}.
This opens the way to a defense that does not merely ask ``does this message look
malicious'', but ``is it consistent with what the organization knows about itself''.
