% ============================================================
% Methodology — verification procedure (target: Computers & Security)
% Procedure, variables, metrics, controls. NO concrete instantiation
% values (those live in the Case Study, Sec.~\ref{sec:results}).
% ============================================================
\section{Methodology}
\label{sec:method}

Having fixed the model (Sec.~\ref{sec:model}), we now describe how we test it: the procedure that turns the two
invariants into detectors, the protocol that guards against inflated results, and the metrics against which the
definition of effectiveness (Sec.~\ref{sec:effectiveness}) is checked. Throughout we keep the description at the
level of a procedure with free parameters; the concrete values that instantiate it (the population, the number of
layers and time buckets, the cascade parameters, the hyperparameters) are stated once, in the Case Study
(Sec.~\ref{sec:results}).

\subsection{Overview of the verification procedure}
\label{sec:m-overview}
The procedure has four steps that mirror the model. (1)~We instantiate the domain knowledge graph $\mathcal{G}$
(Def.~\ref{def:dkg}) on a population, deriving the relation layers $\mathcal{L}$ and the temporal rhythm $\tau$.
(2)~We build the two detectors, one per invariant (Sec.~\ref{sec:invariants}), as off-the-shelf learning
instruments over graph-position features. (3)~We evaluate them under the leak-aware protocol below, so that any
advantage due to leakage or capacity is destroyed by a control. (4)~We check transfer to real topologies and read
off susceptibility (Def.~\ref{def:susc}). Each step is detailed next.

\subsection{Leak-aware evaluation protocol}
\label{sec:eval}
Rigorous evaluation here is a contribution, not a formality, because graph detectors are easy to inflate. The
protocol enforces four requirements. A matched design makes legitimate traffic and attack edges share the same
distribution of endpoints and degrees, so they differ only by the relational-temporal pattern and the
degree/identity confounder is removed \citep{arp2022dos}. Leak control adds a temporal split of the rhythm
estimate, an audit of the fraction of out-of-rhythm mail, and a control that shuffles temporal order, each of
which zeroes the signal if it is an artifact \citep{pendlebury2019tesseract}. Operational metrics report
sensitivity at a low FPR rather than headline AUC, with explicit base-rate arithmetic. A predictivity probe then
asks whether the ranking of methods established on one topology transfers to another (real Enron, SNAP graphs).
Together these give the working definition of effectiveness of Sec.~\ref{sec:effectiveness}: a method works when
its advantage is causally relational-temporal, vanishing under controls, and transferable beyond a single corpus.

\subsection{Detectors and causal controls}
\label{sec:m-detectors}
This subsection turns the two invariants into concrete detectors and states the controls that isolate their
signal. The design principle is uniform: features encode graph position and timing, never identity, so every
detector is inductive and applies to accounts unseen in training.

\smallskip\noindent\textbf{Invariant~I (cross-layer).} For a sender--victim pair we compare three feature sets of
increasing structure, contacts-only, static multiplex (contacts $+$ OSINT layers), and full (adding temporal
consistency), with a gradient-boosting classifier. The causal control shuffles layer membership and rhythm: if
the signal is genuine cross-layer consistency, the control must collapse the advantage to chance.

\smallskip\noindent\textbf{Invariant~II (cascade).} The second invariant operates on the contacts layer
$G=(V,E)$ with time divided into $T$ buckets. An attack is a cascade: a patient-zero $s_0$ sends malicious mail
to contacts; each recipient becomes a carrier with probability $p_\mathrm{inf}$ and propagates to depth $H$
(Fig.~\ref{fig:cascade}). A single event $u\!\to\!v$ is locally indistinguishable from legitimate mail; the
signal is the cascade context, since within the wave's window many of $v$'s neighbors receive mail and $u$ was
hit just before. We compare four detectors of increasing temporal awareness: a 1-hop tabular model (a lower
bound), a hand-crafted cascade context (1-hop plus burst, freshness, and wave density), a static GNN (topology,
no time), and a temporal GNN with node memory.

\smallskip\noindent\emph{Intuition before the equations.} The two equations below differ in one thing only:
\emph{whether the model may look forward in time}. The static score (Eq.~\eqref{eq:static}) summarizes a node by
its neighbors in the whole aggregated graph, seeing who is connected to whom but not in what order. The temporal
score (Eq.~\eqref{eq:temp-score}) instead keeps a running memory $m_v^{(b)}$ that is updated bucket by bucket and
read one bucket before the event it scores, so it can only use the past. That past-only reading is what lets the
receiving-side wave become a signal, and it is also what a time-shuffle destroys, so the control below measures
the temporal signal and nothing else.

\smallskip\noindent Formally, the static GNN builds embeddings from the aggregated graph with a normalized
adjacency $\hat{A}$ and features $X$,
\begin{equation}
H = \hat{A}\,\sigma\!\big(\hat{A}\,X\,W_1\big)\,W_2,
\qquad
\mathrm{score}_{\mathrm{stat}}(s,v) = g\!\big([\,H_s,\,H_v,\,H_s\!\odot\!H_v,\,x_{sv}\,]\big),
\label{eq:static}
\end{equation}
whereas the temporal GNN maintains a memory $m_v^{(b)}$ and scores an event of bucket $b$ using the state from
before its update, which enforces causality and avoids leakage,
\begin{align}
\mathrm{score}_{\mathrm{temp}}(s,v,b) &= g\!\big([\,m_s^{(b-1)},\,m_v^{(b-1)},\,m_s^{(b-1)}\!\odot\!m_v^{(b-1)},\,x_{sv}\,]\big),
\label{eq:temp-score}\\
m_v^{(b)} &= \mathrm{GRU}\!\Big(\tfrac{1}{d_v^{(b)}}\!\!\sum_{u:\,(u\to v)\in b}\!\!\sigma\!\big(W_m\,m_u^{(b-1)}\big),\; m_v^{(b-1)}\Big).
\label{eq:temp-mem}
\end{align}

\emph{Why the shuffle control is exact.} Since $\mathrm{score}_{\mathrm{stat}}$ is permutation-invariant in time
while $\mathrm{score}_{\mathrm{temp}}$ is not, the drop in $\mathrm{AUC}_{\mathrm{temp}}$ after time shuffling
measures the amount of temporal information the model used. This is the formal justification of the temporal
control.

\smallskip\noindent\textbf{Baselines for compromises.} Beyond the GNN we implement two strong practical methods.
A volumetric baseline in the spirit of COMPA \citep{egele2013compa,stringhini2015thataint} flags a sudden spike
in a sender's volume within a short window (an \texttt{IsolationForest}), and a path-based baseline in the spirit
of Hopper \citep{ho2021hopper} looks for a causal precursor, momentum, and a rare target. We implement their
structural-temporal core rather than the full systems, which is why we label them ``volumetric'' and
``path-based''. Hyperparameters of all detectors are fixed uniformly, without intensive tuning, so that no
detector enjoys a tuning advantage; the concrete settings are given in the Case Study (Tab.~\ref{tab:hparams}).

\begin{figure}[t]\centering
\resizebox{0.92\columnwidth}{!}{%
\begin{tikzpicture}[>=Latex, every node/.style={font=\small},
  acc/.style={circle, draw, minimum size=7mm, inner sep=1pt},
  inf/.style={acc, fill=red!18, draw=red!70!black},
  ben/.style={acc, fill=blue!8},
  mal/.style={-{Latex[length=2mm]}, red!70!black, thick}]
  \node[inf] (p) at (0,0) {$s_0$};
  \node[ben] (a) at (2.4,0.9) {$A$};
  \node[ben] (b) at (2.4,-0.9) {$B$};
  \draw[mal] (p)--(a); \draw[mal] (p)--(b);
  \node[ben] (c) at (4.8,1.6) {$C$};
  \node[ben] (d) at (4.8,0.4) {$D$};
  \node[ben] (e) at (4.8,-0.9) {$E$};
  \draw[mal] (a)--(c); \draw[mal] (a)--(d); \draw[mal] (b)--(e);
  \node[ben] (f) at (7.0,1.0) {$F$};
  \draw[mal] (c)--(f);
  \node[gray] at (0,-2.1) {bucket $t_0$};
  \node[gray] at (3.6,-2.1) {$t_1$};
  \node[gray] at (5.9,-2.1) {$t_2$};
  \draw[gray, dashed] (1.2,-1.8) -- (1.2,2.1);
  \draw[gray, dashed] (3.6,-1.8) -- (3.6,2.1);
  \draw[gray, dashed] (5.9,-1.8) -- (5.9,2.1);
  \node[red!70!black, align=center, font=\scriptsize] at (3.6,2.7)
    {edge $A\!\to\!C$ \emph{locally} = ordinary mail from a known contact;\\
     signal = \emph{synchronized burst} of the wave and sender \emph{freshness}};
\end{tikzpicture}}
\caption[Lateral phishing as a temporal cascade]{Lateral phishing as a temporal cascade: a carrier ($s_0$)
infects contacts, those --- their own, and the wave propagates across the graph. Detection requires \emph{multi-hop,
temporal} propagation context.}
\label{fig:cascade}
\end{figure}

\subsection{Event model and data split}
\label{sec:m-events}
An event is a pair (sender $s$, victim $v$) with a bucket $b$, labeled legitimate or attack, with two-sided noise
so that the classes are not perfectly separable. Legitimate events are of three kinds: known (a contact, in
rhythm), new-OSINT (outside contacts but on an OSINT layer), and cold (outside all layers, an irreducible source
of false positives). Attacks are either BEC from outside the graph (outside everything and off-rhythm) or a
compromised account (a known contact off-rhythm). We use two split regimes, victim-wise (disjoint victim sets) and
organization-wise (test organizations disjoint from training ones, an inductive validation). Headline claims are
hardened over many seeds; validations on real graphs use fewer seeds, since variance is lower on a fixed
structure, with a Wilcoxon test where significance matters. The exact seed counts are reported with each result.

\subsection{Metrics, base rate, and threats to validity}
\label{sec:m-metrics}
The headline metric is sensitivity at a fixed low FPR, the operational working point,
$\text{sensitivity}@\mathrm{FPR} = \max\{\mathrm{TPR}(\theta):\mathrm{FPR}(\theta)\le\alpha\}$; AUC is reported
alongside it. Real systems operate at a far stricter FPR ($<\!4$~FP/million, \citep{ho2019detecting}), so the FPR
we report is a liberal lower bound rather than a deployment threshold; at a realistic prevalence, precision
follows from Bayes' theorem, which we make explicit in the Case Study (Sec.~\ref{sec:res-rigor}). The causal
controls are layer/rhythm shuffling for Invariant~I and temporal-order shuffling for Invariant~II: if the signal
is structural-temporal, the control must destroy the advantage.

\smallskip\noindent\textbf{Threats to validity.} The internal threats are layer redundancy, for which we measure
Jaccard before any gain claim, and label leakage, which we counter by separating features from the hidden
infection state and letting the shuffle control isolate the mechanism. The external threat is that the population
is synthetic, so we validate the mechanism on real Enron and two SNAP graphs; the synthetic data lets us control
coverage and fabrication at the cost of absolute effects. The construct threat is addressed by preferring
sensitivity at low FPR to headline AUC, in line with \citep{arp2022dos,pendlebury2019tesseract}.
