%% English version in the Elsevier CAS template (single-column) — target: Computers & Security
\documentclass[a4paper,fleqn]{cas-sc}

\usepackage[authoryear,longnamesfirst]{natbib}

%% Content packages (CAS already loads: graphicx, amsmath, booktabs, multirow, array, hyperref, xcolor)
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{siunitx}\sisetup{output-decimal-marker={.}, group-separator={}, per-mode=symbol}
\usepackage{tikz}\usetikzlibrary{positioning, fit, backgrounds, arrows.meta}
\usepackage{enumitem}
\graphicspath{{figures/}}

%% Theorem-like environments (used in the Model section)
\newtheorem{definition}{Definition}

\begin{document}
\let\WriteBookmarks\relax

%% Running head
\shorttitle{Spear phishing and the domain knowledge graph}
\shortauthors{K. Warpechowski, B. Ksi{\k{e}}{\.z}opolski}

%% Main title
\title[mode=title]{Spear Phishing as a Violation of the Domain Knowledge Graph:
Cascade Dynamics and Leak-Aware Evaluation}

%% First (corresponding) author, affiliations 1 and 2
\author[1,2]{Kamil Warpechowski}
\cormark[1]
\ead{kamil.warpechowski@kozminski.edu.pl}
\credit{Conceptualization, Methodology, Software, Experiments, Writing -- original draft}

\affiliation[1]{organization={Kozminski University},
            city={Warsaw},
            country={Poland}}
\affiliation[2]{organization={NASK National Research Institute},
            city={Warsaw},
            country={Poland}}

%% Second author
\author[1]{Bogdan Ksi{\k{e}}{\.z}opolski}
\ead{bogdan.ksiezopolski@kozminski.edu.pl}
\credit{Supervision, Conceptualization, Writing -- review and editing}

\cortext[1]{Corresponding author}

%% Abstract (inlined -- CAS abstract environment does not expand \input reliably)
\begin{abstract}
Modern email filters detect mass phishing with high accuracy ($\sim$98\%) from content and reputation, yet the
residual traffic, spear phishing and Business Email Compromise (BEC), is the most costly, and nearly undetectable by content and reputation alone. We argue that this is structural: a targeted attack succeeds by respecting the victim's domain
knowledge, so the signal lies not in the content but in its inconsistency with that knowledge and in the attack's
propagation dynamics. We model the organizational domain knowledge graph as a multiplex of relations and study
two invariants of its violation.

The main result is cascade dynamics: a compromised account propagates off-rhythm, leaving a
receiving-side wave signature that persists under a stealthy attacker, where volume-based detection drops below
the random baseline, and that replicates across three real graphs (Enron, SNAP) with large effect sizes. We validate the complementary cross-layer inconsistency invariant synthetically, as a proof of soundness rather
than a deployment result: its OSINT multiplex stays synthetic even on real graphs and a temporal layer dominates the signal, which
we report rather than overstate.

The central, detector-independent contribution is methodological: a leak-aware evaluation with shuffle controls,
a predictivity probe for cross-topology transfer, and explicit base-rate accounting that locates the operational
limits, which remain far from deployment thresholds. An organizational susceptibility metric shows that exposure
is governed by relation rather than rank. The contribution is thus a concept, a measurement methodology, and a
triage signal, with learning models as off-the-shelf instruments: spear phishing becomes measurable, though not
yet operationally solved.
\end{abstract}

%% Research highlights (each <= 85 characters incl. spaces)
\begin{highlights}
\item Cascade dynamics flag lateral phishing from a receiving-side wave signature.
\item The cascade signal survives stealthy attacks where volume detection collapses.
\item Cross-layer inconsistency is validated synthetically as a proof of soundness.
\item Leak-aware evaluation and a predictivity probe bound detection effectiveness.
\item Susceptibility is governed by relation, not organizational rank or seniority.
\end{highlights}

%% Keywords
\begin{keywords}
spear phishing \sep Business Email Compromise \sep domain knowledge graph \sep
cross-layer inconsistency \sep lateral phishing \sep organizational susceptibility
\sep leak-aware evaluation \sep graph neural networks
\end{keywords}

\maketitle

%% ==== Main text (English sections) ====
\input{en/introduction}
\input{en/related_work}
\input{en/model}
\input{en/methodology}
\input{en/results}
\input{en/discussion}
\input{en/conclusion}

%% ==== Back matter: required declarations (C&S Guide for Authors) ====
\section*{CRediT authorship contribution statement}
\textbf{Kamil Warpechowski:} Conceptualization, Methodology, Software, Investigation, Validation,
Writing -- original draft. \textbf{Bogdan Ksi{\k{e}}{\.z}opolski:} Supervision, Conceptualization,
Writing -- review \& editing.

\section*{Declaration of competing interests}
The authors declare that they have no known competing financial interests or personal relationships
that could have appeared to influence the work reported in this paper.

\section*{Funding}
% TODO: confirm with co-author/NASK whether a specific grant supported this work; replace if so.
This research did not receive any specific grant from funding agencies in the public, commercial,
or not-for-profit sectors.

\section*{Data availability}
% TODO before submission: replace XXXXXXX with the DOI reserved on Zenodo
% (New upload -> "Reserve DOI"); the same DOI is recorded in CITATION.cff and .zenodo.json.
The reproducibility package --- the synthetic population generator, all experiment and analysis code,
the generated population of $150$~personas, and the computed results from which the tables and figures
are produced --- is openly archived on Zenodo (DOI:~\texttt{10.5281/zenodo.XXXXXXX}). The real
communication graphs used for validation are public: the Enron email corpus \citep{klimt2004enron}
and the SNAP datasets \texttt{email-Eu-core} and \texttt{CollegeMsg} \citep{leskovec2014snap}.

\section*{Declaration of generative AI and AI-assisted technologies in the manuscript preparation process}
During the preparation of this work the authors used Claude (Anthropic) in order to translate the
manuscript from Polish into English and to copy-edit the prose. After using this tool, the authors
reviewed and edited the content as needed and take full responsibility for the content of the
published article.

%% Bibliography in Elsevier CAS author-year style
\bibliographystyle{cas-model2-names}
\bibliography{bibliography}

\end{document}
